% discussion.tex -- Discussion and Conclusion (merged to save space)

The experiments validated the core claim: Langevin dynamics on the modern Hopfield energy converts deterministic attention into a principled stochastic sampler governed by a single temperature parameter. The synthetic experiments confirmed a smooth phase transition between retrieval and generation (Fig.~\ref{fig:experiments}a), convergence to the correct Boltzmann target (Fig.~\ref{fig:experiments}b), and graceful degradation with load ratio (Fig.~\ref{fig:phase-diagram}). The strongest result came from protein sequences (Table~\ref{tab:protein-main}), where the training-free score function preserved family-level fidelity at position-specific and co-evolutionary levels that the VAE could not maintain from $K{=}68$ examples. On MNIST, SA led the best learned baseline by $2.6{\times}$ on novelty and $2.0{\times}$ on diversity, while DDPM failed across all memory sizes tested.

A deliberate choice shapes our experimental design: we evaluate at memory sizes $K$ from 68 to 3{,}500, the low-to-moderate data regime where learned generative models are most constrained. This is the method's natural operating point. Score-based and diffusion models require enough data to learn an accurate score function; when $K$ is small, this learning problem is ill-posed, as DDPM's failure illustrates. Stochastic attention sidesteps this bottleneck: the score $\nabla\log p_\beta = \beta(\mathbf{T}-\boldsymbol{\xi})$ is exact for any $K$. As $K$ grows, learned models will eventually close the gap; the scaling study (Appendix~\ref{app:scaling}) begins to map this crossover.

In summary, stochastic attention reuses the same primitive operations as a standard attention head and requires no learned score network, no training loop, and no contrastive objective. The Hopfield energy's analytic structure yields convergence guarantees in the convex regime (Corollary~\ref{cor:convergence}) and geometric ergodicity for all $\beta>0$; at high $\beta$ the energy becomes multimodal and mixing bounds remain open, but the multi-chain protocol provides a practical workaround. The entropy inflection (Proposition~\ref{prop:entropy-derivative}) gives a data-dependent criterion for the retrieval--generation boundary; on the RRM protein family it correctly predicted a lower $\beta^*{=}3.85$ (half the random-pattern value $\sqrt{d}{=}7.68$), reflecting the increased effective similarity variance from conserved residues. Because $\mathbf{X}$ can be the key matrix of any pretrained attention layer, the method is a candidate for zero-shot stochastic decoding in retrieval-augmented generation and in-context learning settings; we leave this extension, along with integration with learned multi-head projections, to future work. Concurrent work demonstrates that the framework extends to structural validation across eight protein families with ESMFold and AlphaFold2 predictions~\cite{varnerTrainingFreeProtein2026}, and that adding a scalar bias to the attention logits enables conditioned generation toward functional subsets without retraining~\cite{varnerConditioningProtein2026}, suggesting that the Hopfield energy landscape supports a broader class of generation tasks than explored here.

\paragraph{Limitations.}
ULA discretization bias scales with $\alpha$; MALA (Appendix~\ref{app:mala-algorithm}) is preferred at larger step sizes, though at $\alpha{=}0.01$ the $99.2\%$ acceptance rate confirms negligible bias. At high temperature ($\beta{=}200$), SA samples are structured interpolations rather than crisp novel images; higher-fidelity generation requires operating closer to the phase-transition band. Inter-basin mixing at high $\beta$ grows exponentially with barrier height; quantitative scaling laws remain to be established. Each step costs $\mathcal{O}(dK)$, identical to a single attention head. The method requires $\mathbf{X}$ to be known and fixed; streaming or latent memory settings would require extension. A toy multi-head extension (Appendix~\ref{app:multihead}) shows the mechanism generalizes, but integration with learned projections and full transformers is future work.
