% introduction.tex -- Introduction
% Three-paragraph structure: (1) problem, (2) landscape, (3) contribution.

Attention is the central computational primitive of modern deep learning~\cite{vaswaniAttentionAllYou2017}. Given a query, it computes a softmax-weighted average of stored values, a powerful operation but a fundamentally \emph{deterministic} one. The same query always produces the same output. Attention retrieves; it does not generate. Yet generation from a structured memory is precisely what many downstream tasks require: producing novel but plausible continuations, interpolating between stored prototypes, or exploring the space of patterns consistent with partial evidence. This limitation is sharpest in the low-data regime---small protein families, rare event catalogs, specialized image collections---where the memory bank contains tens to hundreds of examples, far too few to train a score network or diffusion model, yet enough to define a structured energy landscape. A natural question arises: \emph{can the attention mechanism itself be made stochastic, in a principled way, so that it samples from the space of memories rather than merely returning their weighted average?}

The ingredients for an answer already exist, though they have not been assembled. On one side, a line of work on associative memory, from Hopfield networks~\cite{hopfieldNeuralNetworksPhysical1982} through dense associative memories~\cite{krotovDenseAssociativeMemory2016,krotovLargeAssociativeMemory2021} to the modern continuous formulation of Ramsauer et al.~\cite{ramsauerHopfieldNetworksAll2021}, has revealed that each attention head implicitly performs gradient descent on a smooth, confining energy whose minima are stored patterns. On the other side, a mature theory of Langevin dynamics~\cite{robertsTweedieExponentialConvergence1996,wellingBayesianLearningStochastic2011,durmusNonasymptoticAnalysisUnadjusted2017} shows how to convert \emph{any} such energy into a sampler for the corresponding Boltzmann distribution: one simply adds calibrated noise to the gradient update. Meanwhile, energy-based models~\cite{lecunTutorialEnergyBased2006,songHowTrainEnergyBased2021} and score-based diffusion methods~\cite{songGenerativeModelingEstimating2019,songScoreBasedGenerativeModeling2021} have demonstrated the power of Langevin-type sampling for generation, but rely on black-box neural networks whose scores must be learned. The Energy Transformer~\cite{hooverEnergyTransformer2024} brings Hopfield energies into a deep architecture, yet remains a discriminative model that descends the energy rather than sampling from it. The classical retrieval-generation duality (Hopfield networks retrieve, Boltzmann machines sample, and both are defined by the same quadratic interaction energy over binary states~\cite{ackeleyLearningAlgorithmBoltzmann1985}) has not been lifted to the modern continuous setting.

In this paper, we close that gap. We show that applying the unadjusted Langevin algorithm to the modern Hopfield energy yields a \emph{stochastic attention} update: a single iteration comprising a contraction toward the origin, a softmax attention pull toward stored memories, and an isotropic Gaussian perturbation whose magnitude is governed by the temperature. The inverse temperature $\beta$ interpolates continuously between exact retrieval ($\beta\to\infty$) and open-ended exploration ($\beta\to 0$), providing a principled generation mechanism that requires no learned score network, no training loop, and no contrastive objective. Because the gradient of the Hopfield energy is exactly the identity minus the attention map, every step of the sampler is computed by the same primitive operations as a standard attention head. In particular, $\mathbf{X}$ can be the key matrix of any pretrained attention layer, making stochastic attention a zero-shot stochastic decoding layer. When the attention head uses learned $Q$, $K$, $V$ projections operating on hidden states rather than raw patterns, the connection requires projecting through these learned transformations; we leave this extension, along with applications to retrieval-augmented generation and in-context learning, to future work. The energy's analytic structure (infinite differentiability, Lipschitz gradients, and a quadratic confining bound) delivers convergence guarantees that generic energy-based models cannot offer without additional assumptions. In practice, the same algorithm realizes two qualitatively distinct operating regimes: at high $\beta$ it performs \emph{structured retrieval}, generating outputs that closely resemble stored patterns; once $\beta$ falls below a critical signal-to-noise ratio $\mathrm{SNR}^*$ whose scaling is derived from the entropy inflection of the attention distribution (Section~4), it performs \emph{genuine generation}, producing novel outputs that outperform learned baselines on both novelty and diversity while preserving domain-level fidelity (e.g., $6.9{\times}$ lower amino acid composition divergence than a VAE on protein sequences). The individual ingredients, the Hopfield energy gradient~\cite{ramsauerHopfieldNetworksAll2021} and the Langevin algorithm~\cite{robertsTweedieExponentialConvergence1996}, are known. The contribution lies in their synthesis and in the analysis it enables, which is inaccessible from either ingredient alone. First, because the Hopfield energy gradient is exactly the identity minus the attention map, the score function is available in closed form at the cost of a single attention evaluation; this eliminates the score matching or contrastive divergence required by standard energy-based and score-based generative models, and converts any pretrained attention head into a sampler with no additional training. Second, Ramsauer et al.\ studied the energy only for retrieval ($\beta\to\infty$) and the Langevin literature does not know about attention; by analyzing the energy \emph{as a sampling target} we characterize the full temperature spectrum, and the entropy inflection analysis (Proposition~\ref{prop:entropy-derivative}) identifies the critical $\beta^*$ at which the sampler transitions from retrieval to generation, with a scaling law $\beta^*\sim\sqrt{d}$ that yields a dimension-dependent SNR criterion for temperature selection. Third, the smoothness, dissipativity, and condition number of the Hopfield energy as a function of $\beta$ and $K/d$ had not been characterized for the purpose of sampling guarantees; we provide these bounds (Proposition~\ref{prop:regularity}) and show where they apply and where they break down. Section~2 surveys related work; Section~3 reviews the necessary background on attention, modern Hopfield networks, and Langevin dynamics; Section~4 derives the stochastic attention update, presents the algorithm, and characterizes its properties; Section~5 validates both regimes experimentally across five domains.
