% relatedwork.tex -- Related Work

Our work lies at the intersection of associative memory, energy-based modeling, and Langevin sampling. We review each strand and identify where existing approaches stop short of the connection we propose.

\paragraph{Hopfield networks and associative memory.}
Hopfield~\cite{hopfieldNeuralNetworksPhysical1982} introduced a recurrent network of $N$ binary neurons whose asynchronous dynamics minimize the quadratic energy $E(\mathbf{s})=-\frac{1}{2}\mathbf{s}^{\top}\mathbf{W}\mathbf{s}$, with $\mathbf{W}=\frac{1}{N}\sum_{i}\mathbf{m}_i\mathbf{m}_i^{\top}$ the Hebbian weight matrix. Stored patterns correspond to local minima of $E$, so retrieval proceeds by gradient descent from an initial probe. When the $K$ patterns are uncorrelated random binary vectors, Amit et al.~\cite{amitStoringInfiniteNumbers1985} showed via a replica-symmetric mean-field analysis that reliable retrieval is possible only when the load ratio $K/N < \alpha_c \approx 0.138$ (where $\alpha_c$ denotes the critical load ratio, not to be confused with the step size $\alpha$ used later); beyond this threshold the energy landscape becomes dominated by spurious mixtures. Folli et al.~\cite{folliMaximumStorageCapacity2017} later gave exact finite-$N$ expressions for the bit-error probability, confirming the $\alpha_c$ transition and showing that the number of retrieval errors grows with $K/N$ as cross-talk noise overwhelms the coherent signal.

The modern Hopfield network program, initiated by Krotov and Hopfield~\cite{krotovDenseAssociativeMemory2016}, addresses these shortcomings by replacing the quadratic interaction with higher-order (polynomial or exponential) interaction functions. Dense associative memories with degree-$n$ polynomial energy store on the order of $N^{n-1}$ patterns, breaking the linear capacity barrier by trading pairwise for higher-order synaptic interactions, and the resulting retrieval dynamics map onto deep networks with rectified polynomial activations. At the top of this hierarchy sits the log-sum-exp (lse) energy, which achieves storage capacity that grows \emph{exponentially} in the pattern dimension $N$~\cite{ramsauerHopfieldNetworksAll2021}. Ramsauer et al.~\cite{ramsauerHopfieldNetworksAll2021} proved that the continuous-state retrieval map associated with the lse energy is equivalent to transformer-style attention: each attention head performs one step of gradient descent on the energy, and the fixed points of the dynamics (global, metastable, or single-pattern) correspond to different retrieval regimes. All of these works, however, focus exclusively on \emph{retrieval}, that is, the convergence of iterates to stored patterns or their convex combinations. Sampling from the Boltzmann distribution defined by the same energy, which would yield a generative mechanism, is not explored. Our work fills this gap by applying Langevin dynamics to the modern Hopfield energy, converting the deterministic retrieval map into a stochastic sampler.

The idea that a single energy function supports both retrieval and generation dates to the Boltzmann machine~\cite{ackeleyLearningAlgorithmBoltzmann1985}, which shares the quadratic energy of the Hopfield network but samples from it via Gibbs dynamics instead of minimizing it. Sejnowski~\cite{sejnowskiHigherOrderBoltzmann1986} extended this to higher-order interactions, corresponding to the polynomial energies later studied by Krotov and Hopfield~\cite{krotovDenseAssociativeMemory2016}. However, these classical models operate on binary states with discrete Gibbs sampling; our formulation inherits the same duality in the continuous, modern Hopfield setting where Langevin dynamics scales naturally to high dimensions.

\paragraph{Energy-based, score-based and diffusion models.}
Energy-based models (EBMs) define a probability distribution $p(\mathbf{x})\propto\exp(-E(\mathbf{x}))$ for an energy function $E$ parameterized by a neural network~\cite{lecunTutorialEnergyBased2006}. Training typically proceeds by contrastive divergence~\cite{hintonTrainingProductsExperts2002}, score matching~\cite{hyvarinen2005estimation}, or noise contrastive estimation~\cite{gutmannHyvaerinenNoiseContrastive2010,songHowTrainEnergyBased2021}. The energy in a generic EBM is a black-box neural network, so its gradient (the score function) must be estimated or approximated. By contrast, the modern Hopfield energy has a \emph{closed-form} gradient (identity minus softmax attention) requiring no auxiliary score network. Moreover, $\mathbf{X}$ is given as data rather than learned, so no training loop is needed, and the log-sum-exp structure provides analytic guarantees (smoothness, Lipschitz gradients, quadratic confinement) that generic EBMs achieve only with careful architectural design. The Energy Transformer~\cite{hooverEnergyTransformer2024} shows that Hopfield-type energies improve discriminative tasks (graph anomaly detection, image completion) but does not sample from the Boltzmann distribution; our work uses the same energy class for generation.

Song and Ermon~\cite{songGenerativeModelingEstimating2019} introduced score-based generative modeling, in which a neural network is trained to approximate $\nabla\log p(\mathbf{x})$ at multiple noise levels, and samples are drawn by annealed Langevin dynamics. Song et al.~\cite{songScoreBasedGenerativeModeling2021} unified this framework with diffusion probabilistic models~\cite{hoDenoisingDiffusionProbabilistic2020,sohlDicksteinDeepUnsupervisedLearning2015} by showing that both arise from forward and reverse-time stochastic differential equations. These methods achieve remarkable sample quality but rely on a learned score network trained on data; our score function is \emph{exact} and \emph{analytic}, $\nabla\log p_\beta(\boldsymbol{\xi}) = \beta(\mathbf{T}(\boldsymbol{\xi})-\boldsymbol{\xi})$, requiring no training, at the cost of being restricted to the Boltzmann distribution of a fixed memory $\mathbf{X}$, a restriction that permits convergence guarantees unavailable in the learned-score setting.

Normalizing flows~\cite{rezendeMohamed2015} construct flexible generative models via learned invertible transformations; like diffusion models, they require training data and a learned bijection, and do not expose a closed-form score function over a fixed memory. Several works have introduced stochasticity into the attention mechanism itself: Deng et al.~\cite{dengLatentAlignment2018} replace the softmax with a latent categorical or Gaussian variable, trained via variational inference, to model alignment uncertainty in sequence-to-sequence tasks. These approaches treat attention weights as random variables to be inferred, whereas our stochasticity arises from Langevin dynamics on a fixed Hopfield energy: the temperature parameter $1/\beta$ controls the noise level, no variational objective is optimized, and the goal is memory-based generation rather than alignment uncertainty.