\documentclass{article}

% Pass numbered, compressed citations to natbib
\PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2026

% The authors should use one of these tracks.
% Before accepting by the NeurIPS conference, select one of the options below.
% 0. "default" for submission
%  \usepackage{neurips_2026}
\usepackage[preprint]{neurips_2026}
% the "default" option is equal to the "main" option, which is used for the Main Track with double-blind reviewing.
% 1. "main" option is used for the Main Track
%  \usepackage[main]{neurips_2026}
% 2. "position" option is used for the Position Paper Track
%  \usepackage[position]{neurips_2026}
% 3. "dandb" option is used for the Datasets & Benchmarks Track
 % \usepackage[dandb]{neurips_2026}
% 4. "creativeai" option is used for the Creative AI Track
%  \usepackage[creativeai]{neurips_2026}
% 5. "sglblindworkshop" option is used for the Workshop with single-blind reviewing
 % \usepackage[sglblindworkshop]{neurips_2026}
% 6. "dblblindworkshop" option is used for the Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop]{neurips_2026}

% After being accepted, the authors should add "final" behind the track to compile a camera-ready version.
% 1. Main Track
 % \usepackage[main, final]{neurips_2026}
% 2. Position Paper Track
%  \usepackage[position, final]{neurips_2026}
% 3. Datasets & Benchmarks Track
 % \usepackage[dandb, final]{neurips_2026}
% 4. Creative AI Track
%  \usepackage[creativeai, final]{neurips_2026}
% 5. Workshop with single-blind reviewing
%  \usepackage[sglblindworkshop, final]{neurips_2026}
% 6. Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop, final]{neurips_2026}
% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote.
% For workshops (5., 6.), the authors should add the name of the workshop, "\workshoptitle" command is used to set the workshop title.
% \workshoptitle{WORKSHOP TITLE}

% "preprint" option is used for arXiv or other preprint submissions
 % \usepackage[preprint]{neurips_2026}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2026}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[colorlinks=true,linkcolor={blue!70!black},citecolor={blue!70!black},urlcolor={blue!70!black},hypertexnames=false]{hyperref}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}       % includegraphics
\usepackage{subcaption}     % subfigures with (a), (b) labels
\usepackage{amsmath,amssymb,amsthm} % math
\usepackage{algorithm}
\usepackage{algorithmic}

\setcounter{topnumber}{1}        % at most 1 float at the top of each page
\setcounter{bottomnumber}{1}     % at most 1 float at the bottom of each page
\renewcommand{\topfraction}{0.5}  % a top float may occupy at most 50% of the page

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote. 
\title{Stochastic Attention via Langevin Dynamics on the Modern Hopfield Energy\thanks{Repository: \url{https://github.com/varnerlab/stochastic-attention-study-paper.git}.}}


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.


\author{%
  Abdulrahman Alswaidan and Jeffrey D. Varner \\
  R.F. Smith School of Chemical and Biomolecular Engineering\\
  Cornell University, Ithaca, NY 14850 \\
  \texttt{\{aa2725, jdv27\}@cornell.edu} \\
}


\begin{document}

\maketitle

\begin{abstract}
Attention heads retrieve: given a query, they return a softmax-weighted average of stored values. We show that this computation is one step of gradient descent on a classical energy function, and that Langevin sampling from the corresponding distribution yields \emph{stochastic attention}: a training-free sampler controlled by a single temperature. Lowering the temperature gives exact retrieval; raising it gives open-ended generation. Because the energy gradient equals the attention map, no score network, training loop, or learned model is required---making the method especially suited to the low-data regime where learned generative models are starved of training signal. We derive a closed-form entropy inflection condition that identifies the retrieval-to-generation transition temperature for any memory geometry, with a scaling law $\beta^*\!\sim\!\sqrt{d}$ for random patterns. We validate on five domains (64 to 4,096 dimensions). On MNIST digit images, stochastic attention is $2.6{\times}$ more novel and $2.0{\times}$ more diverse than the best learned baseline (a VAE trained on the same patterns), while matching a Metropolis-corrected gold standard. On protein sequences from the Pfam RRM family, the generation regime achieves $6.9{\times}$ lower amino acid composition divergence than the VAE (KL $= 0.060$ vs.\ $0.416$) at matched novelty, demonstrating that the training-free score function preserves family-level fidelity that learned models lose. A denoising diffusion baseline (DDPM) fails across all memory sizes tested ($K = 100$ to $3{,}500$), producing samples indistinguishable from isotropic noise. The approach requires no architectural changes to the underlying attention mechanism.
\end{abstract}


\section{Introduction}
\input{sections/introduction.tex}

\section{Related Work}
\input{sections/relatedwork.tex}

\section{Background}
\input{sections/background.tex}

\section{Method}
\input{sections/method.tex}

\section{Experiments}
\input{sections/experiments.tex}

\section{Discussion and Conclusion}
\input{sections/discussion.tex}

\section*{Acknowledgments}
This work grew out of the CHEME-5820 Machine Learning and Artificial Intelligence Methods for Engineers course at Cornell University; the central idea arose while preparing a lecture on Boltzmann machines immediately after covering modern Hopfield networks. Financial market data used in Appendix~\ref{app:finance} were obtained from Polygon.io.

\bibliographystyle{unsrtnat}
\bibliography{References_v1}

\newpage
\appendix
\input{sections/appendix.tex}

\end{document}